\section{Supplemental Material}

\subsection{One-Max Task}
\label{sec:one-max}

We used the one-max task for population size and genealogical inference experiments.
This task selects over binary string of length 100 for the number of 1's set.
Fitness preferred individuals with more 1's over individuals with fewer.
Population size was 200, initialized randomly.
In treatments where population size decreased during an evolutionary run, excess individuals were designated for elimination randomly.
In treatments where population size increased during an evolutionary run,
new population slots were filled through selection over the existing population.

Selection was performed using tournament selection with synchronous generations.
Default tournament size was 2.
Evolutionary runs lasted 100 generations.
Two-point crossover (mating) and bit-flip mutation with per-bit probability 0.05 each generation prior to selection.
Operator choice and parameter selections follows one-max example code from the DEAP package \citep{fortin2012deap}.

\subsection{Population Size Inference Example}
\label{sec:population-size-inference-example}

\input{fig/ne-process-example}

Panel \ref{fig:ne-process-example:differentia} shows magnitudes of fixed differentia across the generational record extracted from a sample specimen at the end of the run.

Population size estimates can be computed at each generation using the maximum likelihood estimator (given in Equation \ref{eqn:popsize_mle}), as shown in Panel \ref{fig:ne-process-example:singleton-est}.
The true population size is annotated as a horizontal blue line.

To generate more robust inference, inference was pooled over rolling sets of 10 fixed differentia.
Population size estimates were again performed using the maximum likelihood estimator given in Equation \ref{eqn:popsize_mle} with confidence interval bounds computed via Equation \ref{eqn:popsize_mle_ci}.
Supplemental Figure \ref{fig:ne-process-example:running-est} plots this running estimation.
Note that some discrepancy is expected between absolute population size (horizontal line) and effective population size $N_e$ (estimated) due to demographic factors.



\subsection{Supplementary Figures}

\input{fig/beta-explain.tex}

\input{fig/species-reconstruction-error}

\input{fig/ne-example-replicates}

\input{fig/ne-detection-matrix}

\input{fig/selection-sensitivity-specificity}

% \input{text/supplement/example.tex}
